\section{Discussion}
\paragraph{Where tree structure helps.}
The results confirm that tree-structured memory provides clear advantages for questions requiring multi-step or temporal reasoning, while flat methods retain an edge on open-domain recall. The 9.0-point F1 gain on multi-hop and 13.8-point gain on temporal questions over Memory~R1 suggest that the hierarchical grouping of facts by speaker, topic, and session enables retrieval paths that flat scoring cannot replicate. Multi-hop questions benefit because related facts (e.g., a person's job and the city where they work) are co-located under a shared ancestor, reducing the retriever's search space. Temporal questions benefit because events are organized chronologically within topic branches, making recency and ordering explicit. At the same time, the dominance of RAG~(BM25) on open-domain questions (38.7 F1) highlights that broad keyword coverage remains valuable when questions do not require structural navigation.

\paragraph{Lexical vs.\ semantic evaluation.}
The consistent F1--J divergence across baselines is informative. Methods that use structured memory (Semantic XPath, our method) tend to achieve higher LLM-judge accuracy relative to their F1, suggesting that tree-guided retrieval surfaces more contextually complete evidence even when the final answer does not lexically match the gold reference. This pattern reinforces the value of evaluating with both lexical and semantic metrics, and suggests that F1 alone may underestimate the advantage of structured retrieval.

\paragraph{Oracle structure caveat.}
The current formulation uses oracle structured data for memory construction, meaning the Memory Manager has not yet been trained end-to-end. This represents an important caveat: the reported gains reflect the ceiling of what tree structure can provide given perfect writing, not what an end-to-end system would achieve. Completing the Memory Manager training loop is the most important next step. We expect some degradation with automatically induced structure, but the ablation study shows that the Retrieve Agent RL contributes the largest share of performance---suggesting that even imperfect tree structure, when combined with learned navigation, should outperform flat alternatives.

\paragraph{Modular training.}
More broadly, the results suggest that structural ambition must be balanced against training stability. The present method uses stable CRUD updates for memory writing, while the Retrieve Agent carries the burden of structure-aware navigation. This separation is cleaner than asking a single policy to both reorganize memory arbitrarily and discover retrieval strategies simultaneously, and is supported by the ablation: disabling Retrieve Agent RL causes the largest performance drop ($-$4.8 F1), confirming that learned tree navigation is the most critical component.
