\section{Problem Definition}

We formalize the setting of a large language model augmented with external memory for long-horizon conversational question answering.

\paragraph{Memory-Augmented Generation.}
Let $\pi_\theta$ denote an LLM parameterized by $\theta$. At each turn $t$, the model receives a user query $x_t \in \mathcal{X}$ and has access to an external memory state $M_t \in \mathcal{M}$, where $\mathcal{M}$ is the space of all possible memory configurations. The model produces a response:
\begin{equation}
y_t \sim \pi_\theta(\cdot \mid x_t, R_t), \qquad R_t = \textsc{Retrieve}(x_t, M_t),
\end{equation}
where $R_t \subseteq M_t$ is a subset of memories retrieved as context. The memory is updated after each turn:
\begin{equation}
M_{t+1} = \textsc{Update}(M_t, x_t, y_t).
\end{equation}
The objective is to learn the retrieval policy $\textsc{Retrieve}$ and the update policy $\textsc{Update}$ such that $\pi_\theta$ maximizes downstream task performance across a sequence of interactions $\mathcal{I} = (x_1, x_2, \ldots, x_T)$.

\paragraph{Flat Memory.}
In prior work~\cg{yan2025memory, jin2025search}, the memory state is a flat, unordered collection of $N$ entries:
\begin{equation}
M_t^{\text{flat}} = \{m_1, m_2, \ldots, m_N\}, \qquad m_i = (\text{key}_i,\; \text{value}_i),
\end{equation}
where each entry $m_i$ is an independent fact. Retrieval is performed by scoring each entry against the query independently, typically via embedding similarity or keyword matching:
\begin{equation}
R_t = \text{Top-}k \left\{ m_i \in M_t^{\text{flat}} : \text{score}(x_t, m_i) \right\}.
\end{equation}
This formulation treats all memories as exchangeable and cannot exploit structural relationships among them.

\paragraph{Tree-Structured Memory.}
We propose to organize memory as a rooted tree $\mathcal{T} = (V, E)$ where vertices $V$ are organized into $L$ levels. The root $v_0$ represents the entire conversation. Internal nodes at level $\ell$ represent progressively finer groupings (e.g., people $\to$ topics $\to$ events), and leaf nodes contain individual facts:
\begin{equation}
M_t^{\text{tree}} = \mathcal{T}_t = (V_t, E_t), \qquad V_t = V_t^{(0)} \cup V_t^{(1)} \cup \cdots \cup V_t^{(L)},
\end{equation}
where $V_t^{(\ell)}$ denotes nodes at level $\ell$ and $E_t$ encodes parent-child relationships. Each node $v \in V_t$ carries a content string $c(v)$ and a summary $s(v)$ that aggregates information from its descendants.

Under tree-structured memory, retrieval becomes a \emph{search planning} problem. Rather than scoring all entries globally, the retriever selects a path through the hierarchy:
\begin{equation}\label{eq:tree_retrieve}
R_t = \textsc{TreeRetrieve}(x_t, \mathcal{T}_t) = \textsc{Execute}\!\big(\mathcal{T}_t,\; z_t\big), \qquad z_t \sim \pi_{\theta_R}(\cdot \mid x_t, \textsc{Schema}(\mathcal{T}_t)),
\end{equation}
where $z_t$ is a search plan specifying which branches to expand and which leaves to read, and $\textsc{Schema}(\mathcal{T}_t)$ is a compact representation of the tree structure (node names and summaries, without full leaf content).

Similarly, updates become a \emph{structural placement} problem. Given a new fact extracted from turn $x_t$, the update policy must decide where in the hierarchy it belongs:
\begin{equation}\label{eq:tree_update}
\mathcal{T}_{t+1} = \textsc{TreeUpdate}(\mathcal{T}_t, x_t) = \textsc{Apply}\!\big(\mathcal{T}_t,\; o_t\big), \qquad o_t \sim \pi_{\theta_M}(\cdot \mid x_t, \mathcal{T}_t),
\end{equation}
where $o_t$ is an update operation that places, merges, or removes facts within the tree.

\paragraph{Learning Objective.}
Both the search plan $z_t$ and the update operation $o_t$ are intermediate decisions whose quality can only be assessed through the correctness of the final answer $y_t$ to a downstream question. Given a dataset of question-answer pairs $\{(q_j, y_j^\star)\}_{j=1}^{N_q}$, we optimize:
\begin{equation}\label{eq:objective}
\max_{\theta_M, \theta_R, \theta_A} \;\; \mathbb{E}_{q,y^\star} \left[ \mathcal{R}(y, y^\star) \right], \qquad y \sim \pi_{\theta_A}(\cdot \mid q, R), \quad R = \textsc{TreeRetrieve}(q, \mathcal{T}),
\end{equation}
where $\mathcal{R}$ is a reward function measuring answer quality and $\mathcal{T}$ is the memory tree constructed by the Memory Manager policy $\pi_{\theta_M}$. This objective couples memory construction, retrieval, and answering through a shared downstream reward signal.
