\section{Related Work}
\paragraph{Memory-Augmented LLMs and RAG.}
Retrieval-augmented generation improves factuality and coverage by accessing external knowledge at inference time~\cg{lewis2020retrieval}. Flat nearest-neighbor memory has likewise been used to improve language-model behavior through direct memory lookup~\cg{khandelwal2020generalization}. In agent settings, memory is often maintained as a chronological stream of observations, reflections, or summaries~\cg{park2023generative}. A recent survey~\cg{zhang2024survey} provides a comprehensive overview of memory mechanisms for LLM-based agents. These approaches demonstrate the value of persistent memory but treat it as a flat collection, without exploiting the event and topic structure inherent in conversation.

\paragraph{Structured and Hierarchical Memory.}
Several methods organize memory hierarchically to improve access. RAPTOR~\cg{sarthi2024raptor} uses recursive abstractive processing to build tree-organized retrieval structures. MemWalker~\cg{chen2023memwalker} navigates memory trees interactively, reading beyond context limits. MemTree~\cg{rezazadeh2024memtree} explores tree-structured memory for conversational agents. Lattice~\cg{lattice2025} introduces LLM-guided hierarchical retrieval over structured memory. Semantic XPath~\cg{semanticxpath2026} generates path-like queries over memory trees. These works show that hierarchical organization can improve memory access, but typically assume the structure already exists or rely on heuristic construction methods. Our work differs in that the tree is induced incrementally from raw conversation and access over it is optimized as a learned policy using downstream reward.

\paragraph{RL for Search and Memory.}
Search-R1 treats external search as a policy decision and optimizes reasoning-search trajectories with reinforcement learning~\cg{jin2025search}. Memory-R1 extends this perspective to persistent memory, training a Memory Manager and an Answer Agent using downstream question-answering reward~\cg{yan2025memory}. These works establish the key principle that information access and memory manipulation should be trained from task outcomes rather than from heuristic rules. However, Search-R1 addresses external search rather than persistent conversational memory, and Memory-R1 operates over flat memory without addressing tree-structured access or updates. Our work inherits this outcome-driven perspective while extending it to tree-structured memory with learned access and update policies.

\paragraph{Gap addressed by this work.}
Prior work treats memory structure, learned access, and learned updates as separate concerns. Hierarchical memory methods~\cg{sarthi2024raptor, chen2023memwalker, rezazadeh2024memtree, lattice2025} assume or heuristically construct structure but do not learn access policies; RL-based methods~\cg{jin2025search, yan2025memory} learn access policies but operate over flat memory. \textsc{Struct Memory-R1} addresses all three within a unified RL framework: it induces tree structure from raw conversation and trains both tree-aware access and update policies from downstream QA reward.
